---
title: "Covid19R Project Data Format"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Covid19R Project Format}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


The `refresh_coronavirus_jhu` function enables to pull the `coronavirus` dataset using the [Covid19R](https://covid19r.github.io/documentation/) project [data format standard](https://covid19r.github.io/documentation/data-format-standard.html). This data format includes the following fields:

* `date` - The date of the observation, using `Date` class
* `location` - The name of the location as provided by John Hopkins raw data
* `location_type` - The type of the location field, either `country` or `state`
* `location_code` - The location code, using the `ios_3166_2` format
* `location_code_type` - The location code type (`ios_3166_2`)
* `data_type` - The case type, `c("recovered_new", "cases_new", "deaths_new" )`
* `value` - The number of cases
* `lat` - The latitude code
* `long` - The longitude code


```{r }
library(coronavirus)

covid19_df <- refresh_coronavirus_jhu()

head(covid19_df)
```

