---
title: "Covid19 Vaccine Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Covid19 Vaccine Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.height=5, fig.width=8, message=FALSE, warning=FALSE,
  collapse = TRUE,
  comment = "#>"
)
```


This vignette provides an overview to the `covid19_vaccine` dataset. 

## Data

Before getting started, please note the following:

* It may be challenging to compare countries as each may use different count methods for the given doses.
* In addition, the dose summary includes multiple types of vaccines, and there is no information if the doses are a single or multiple doses series.

**Data sources:**

-   Vaccine data - <https://github.com/govex/COVID-19/tree/master/data_tables/vaccine_data/global_data>
-   Country/province codes and population - <https://github.com/CSSEGISandData/COVID-19/blob/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv>
-   Continent names and code - DATA HUB, <https://datahub.io/JohnSnowLabs/country-and-continent-codes-list>


### Load the data

Let's start by loading the packages we will use in this vignette:

```{r}
library(coronavirus)
library(dplyr)
library(plotly)
```

We load the dataset from the **coronavirus** package, and use the **dplyr** and **plotly** packages to manipulate and plot the data. Let's load the data:

```{r}
data("covid19_vaccine")

head(covid19_vaccine)
```


The `covid19_vaccine` dataset provides time-series data on the vaccination progress by country or province (if applicable). Likewise, the `coronavirus` dataset, the COVID19 vaccine raw data is collected by Johns Hopkins University Center for Systems Science and Engineering (JHU CCSE). The `covid19_vaccine` data includes the following fields:

- `date` - Data collection date in YYYY-MM-DD format
- `country_region` - Country or region name
- `continent_name` - Continent name
- `continent_code` - Continent code
- `combined_key` - Country and province (if applicable)
- `doses_admin` - Cumulative number of doses administered. When a vaccine requires multiple doses, each one is counted independently
- `people_at_least_one_dose` - Cumulative number of people who received at least one vaccine dose. When the person receives a prescribed second dose, it is not counted twice
- `population` - Country or province population
- `uid` - Country code
- `iso2` - Officially assigned country code identifiers with two-letter
- `iso3` - Officially assigned country code identifiers with three-letter
- `code3` - UN country code
- `fips` - Federal Information Processing Standards code that uniquely identifies counties within the USA
- `lat` - Latitude
- `long` - Longitude


**Note:** The country / province code fields (e.g., `ios2`, `ios3`, etc.) and population were merged with the raw data


### Refresh the data

The data in the package is up-to-date with the last date the package was updated on CRAN. To get the most recent data available on the source (data get updated daily), you can use one of the following methods:

- Use the `update_dataset` which will re-install the package from the package [repository](https://github.com/RamiKrispin/coronavirus). Please see the  vignette for more details 
- Pull the data directly from the [CSV](https://github.com/RamiKrispin/coronavirus/blob/main/csv/covid19_vaccine.csv) file on the repository:

```{r}
library(readr)

vaccine_df <- read_csv(file = "https://raw.githubusercontent.com/RamiKrispin/coronavirus/main/csv/covid19_vaccine.csv",
                       col_types = cols(
                         date = col_date(format = ""),
                         country_region = col_character(),
                         continent_name = col_character(),
                         continent_code = col_character(),
                         combined_key = col_character(),
                         doses_admin = col_number(),
                         people_at_least_one_dose = col_number(),
                         population = col_number(),
                         uid = col_number(),
                         iso2 = col_character(),
                         iso3 = col_character(),
                         code3 = col_number(),
                         fips = col_logical(),
                         lat = col_number(),
                         long = col_number()
                       )) 

# Fixing the continent code field - changing NA to "NA" for North America
vaccine_df$continent_code <- ifelse(is.na(vaccine_df$continent_code) & vaccine_df$continent_name == "North America", "NA", vaccine_df$continent_code)

head(vaccine_df)
```


The plan, in the next version, is to add a similar function to the `refresh_coronavirus_jhu` function to pull the vaccine dataset.

## Analysing the vaccine dataset

The dataset includes two cumulative counts of vaccine doses received: 

- `people_at_least_one_dose` - counts both single and multiple doses of vaccine as one dose
- `doses_admin` - counts each dose received separately

So, for example, if a person received a multi doses vaccine such as the Pfizer or Moderna vaccine, here is the count by each method:

- If the person completed the two doses, the `people_at_least_one_dose` will count it as **one** dose, and the `doses_admin` will count it as **two**
- Similarly, if the person received only the first out of two doses and skipped the second, it will be counted in both variables as **one**

We can assume that the last case is rare; therefore, the `people_at_least_one_dose` could be a good proxy for fully vaccinated people. Let’s see the differences between the two by using the US vaccine data:


```{r}
covid19_vaccine |> 
  filter(country_region == "US") |>
  select(date, doses_admin, total = people_at_least_one_dose) |>
  arrange(date) |>
  plot_ly() |>
  add_lines(x = ~ date,
            y = ~ doses_admin,
            name = "Total Doses") |>
  add_lines(x = ~ date,
            y = ~ total,
            name = "Recevied Vaccine with One/Two Doses") |>
  layout(title = "US Cumulative Number of Doses by Type",
         legend = list(x = 0.05, y = 0.95),
         margin = list(l = 50, r = 50, b = 50, t = 90),
         yaxis = list(title = "Number of Doses"),
         xaxis = list(title = ""))

```


Here are some thoughts about how to leverage this dataset:

- While it is not entirely accurate, as mentioned above, the `people_at_least_one_dose` could provide a good proxy for the total number of fully vaccinated people 
- We can leverage the `doses_admin` column to calculate some stats about the total number of doses provided in a given period or overall. 
- Combining the vaccine and new cases datasets could provide insights into vaccine adoption and effectiveness.

### Country level summary

Before going back and diving into the US data, let's summarize the data by country. Since the data is cumulative, let's take a snapshot of the most recent date in the data and calculate the ratio between the total number of doses and the size of the population:

```{r}
max(covid19_vaccine$date)

df_summary <- covid19_vaccine |>
  filter(date == max(date)) |>
  select(date, country_region, doses_admin, total = people_at_least_one_dose, population, continent_name) |>
  mutate(doses_pop_ratio = doses_admin / population,
         total_pop_ratio = total / population) |>
  arrange(- total)

head(df_summary, 10)

```


It would be interesting to plot the ratio between the number of overall doses received and the size of the population. Let’s filter missing values and plot the ratio:


```{r}
df_summary |> 
  filter(country_region != "World", 
         !is.na(population),
         !is.na(total)) |>
  plot_ly(x = ~ population,
          y = ~ total,
          text = ~ paste("Country: ", country_region, "<br>",
                         "Population: ", population, "<br>",
                         "Total Doses: ", total, "<br>",
                         "Ratio: ", round(total_pop_ratio, 2), 
                         sep = ""),
          type = "scatter",
          mode = "markers") |>
  layout(title = "Total Doses vs. Population",
         margin = list(l = 50, r = 50, b = 60, t = 70),
         yaxis = list(title = "Number of Doses"),
         xaxis = list(title = "Population Size"))
```

As can see, China and India are skewing the plot. We can re-scale the plot with log transformation of the number of doses and population size using Plotly’s built-in log transformation. Let’s also color by continent and add diagonal lines to represent the 1 to 1 relationship between the median and first quartile:

```{r}

# Setting the diagonal lines range
line_start <- 10000
line_end <- 1500 * 10 ^ 6

# Filter the data
d <- df_summary |> 
  filter(country_region != "World", 
         !is.na(population),
         !is.na(total)) 


# Replot it
plot_ly() |>
  add_markers(x = d$population,
              y = d$total,
              text = ~ paste("Country: ", d$country_region, "<br>",
                             "Population: ", d$population, "<br>",
                             "Total Doses: ", d$total, "<br>",
                             "Ratio: ", round(d$total_pop_ratio, 2), 
                             sep = ""),
              color = d$continent_name,
              type = "scatter",
              mode = "markers") |>
  add_lines(x = c(line_start, line_end),
            y = c(line_start, line_end),
            showlegend = FALSE,
            line = list(color = "gray", width = 0.5)) |>
  add_lines(x = c(line_start, line_end),
            y = c(0.5 * line_start, 0.5 * line_end),
            showlegend = FALSE,
            line = list(color = "gray", width = 0.5)) |>
  
  add_lines(x = c(line_start, line_end),
            y = c(0.25 * line_start, 0.25 * line_end),
            showlegend = FALSE,
            line = list(color = "gray", width = 0.5)) |>
  add_annotations(text = "1:1",
                  x = log10(line_end * 1.25),
                  y = log10(line_end * 1.25),
                  showarrow = FALSE,
                  textangle = -25,
                  font = list(size = 8),
                  xref = "x",
                  yref = "y") |>
  add_annotations(text = "1:2",
                  x = log10(line_end * 1.25),
                  y = log10(0.5 * line_end * 1.25),
                  showarrow = FALSE,
                  textangle = -25,
                  font = list(size = 8),
                  xref = "x",
                  yref = "y") |>
  add_annotations(text = "1:4",
                  x = log10(line_end * 1.25),
                  y = log10(0.25 * line_end * 1.25),
                  showarrow = FALSE,
                  textangle = -25,
                  font = list(size = 8),
                  xref = "x",
                  yref = "y") |>
  add_annotations(text = "Source: Johns Hopkins University - Centers for Civic Impact",
                  showarrow = FALSE,
                  xref = "paper",
                  yref = "paper",
                  x = -0.05, y = - 0.33) |>
  layout(title = "Covid19 Vaccine - Total Doses vs. Population Ratio (Log Scale)",
         margin = list(l = 50, r = 50, b = 90, t = 70),
         yaxis = list(title = "Number of Doses",
                      type = "log"),
         xaxis = list(title = "Population Size",
                      type = "log"),
         legend = list(x = 0.75, y = 0.05))
```

Another way to represent the data is by using a box plot to plot the total doses and population ration distribution by continent:


```{r}
plot_ly(d,
        y = ~ total_pop_ratio,
        color = ~ continent_name,
        type = "box", boxpoints = "all", jitter = 0.3,
        pointpos = -1.8
        ) |>
  layout(title = "Distribution of Total Doses Admited vs Population Size Ratio",
         margin = list(l = 50, r = 50, b = 60, t = 60),
         yaxis = list(title = "Total Doses/Population Ratio"),
         xaxis = list(title = "Continent")
    
  )
```

### Plot US cases

Let's now return to the number of doses in the US and plot the daily number of new cases with the daily number of vaccine doses received. We will use the `people_at_least_one_dose` column as a proxy for the number of people vaccinated. First, let's reformat the data from cumulative to daily by subtracting from the cumulative count its first lag:


```{r}
us_vaccine <- covid19_vaccine |> 
  filter(country_region == "US") |>
  select(date, total_doses = people_at_least_one_dose) |>
  mutate(total_doses_lag1 = lag(total_doses, 1),
         daily_doses = total_doses - total_doses_lag1)|>
  select(date, daily_doses) |>
  arrange(date) 


head(us_vaccine)

```

Let's now plot the daily doses:

```{r}
plot_ly(us_vaccine) |>
  add_lines(x = ~ date,
            y = ~ daily_doses)  |>
  layout(title = "US Daily Vaccine Doses Received",
         margin = list(l = 50, r = 50, b = 60, t = 60),
         yaxis = list(title = "Total Doses"),
         xaxis = list(title = ""))
```


We can now add the Covid19 cases by loading the dataset using the `refresh_coronavirus_jhu` function, filter it (to US), and merge it with the `us_vaccine` dataset we created above:

```{r}

covid19_cases <- refresh_coronavirus_jhu()

head(covid19_cases)

us_cases <- covid19_cases |>
  filter(location == "US",
         data_type == "cases_new") |>
  select(date, cases = value)


head(us_cases)
```


Let's now merge the two datasets and plot them:

```{r}
df <- us_vaccine |> 
  left_join(us_cases, by = "date") |>
  select(date, daily_doses, daily_cases = cases) |>
  arrange(date)

head(df)


plot_ly(df) |>
  add_lines(x = ~ date,
            y = ~ daily_cases,
            name = "Daily Cases") |>
  add_lines(x = ~ date,
            y = ~ daily_doses,
            name = "Daily Doses") |>
  layout(title = "US Daily New Cases vs. Doses of Vaccine",
         yaxis = list(title = "New Cases/Doses"),
         xaxis = list(title = ""),
         legend = list(x = 0.75, y = 0.95),
         margin = list(l = 50, r = 50, b = 50, t = 90))
```

