---
title: "Covid19 Vaccine Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Covid19 Vaccine Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.height=5, fig.width=8, message=FALSE, warning=FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

The `covid19_vaccine` dataset provides time-series data on the vaccination progress by country or province (if applicable). Likewise the `coronavirus` dataset, the COVID19 vaccine raw data is collected by Johns Hopkins University Center for Systems Science and Engineering (JHU CCSE). The `covid19_vaccine` data includes the following fields:

-   `country_region` - Country or region name
-   `date` - Data collection date in YYYY-MM-DD format
-   `doses_admin` - Cumulative number of doses administered. When a vaccine requires multiple doses, each one is counted independently
-   `people_partially_vaccinated` - Cumulative number of people who received at least one vaccine dose. When the person receives a prescribed second dose, it is not counted twice
-   `people_fully_vaccinated` - Cumulative number of people who received all prescribed doses necessary to be considered fully vaccinated
-   `report_date_string` - Data report date in YYYY-MM-DD format
-   `uid` - Country code
-   `province_state` - Province or state if applicable
-   `iso2` - Officially assigned country code identifiers with two-letter
-   `iso3` - Officially assigned country code identifiers with three-letter
-   `code3` - UN country code
-   `fips` - Federal Information Processing Standards code that uniquely identifies counties within the USA
-   `lat` - Latitude
-   `long` - Longitude
-   `combined_key` - Country and province (if applicable)
-   `population` - Country or province population
-   `continent_name` - Continent name
-   `continent_code` - Continent code

Data sources:

-   Vaccine data - <https://github.com/govex/COVID-19/tree/master/data_tables/vaccine_data/global_data>
-   Country/province codes and population - <https://github.com/CSSEGISandData/COVID-19/blob/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv>
-   Continent names and code - DATA HUB, <https://datahub.io/JohnSnowLabs/country-and-continent-codes-list>

**Note:** The country / province code fields (e.g., `ios2`, `ios3`, etc.) and population were added to the raw dat

```{r }
library(coronavirus)

data("covid19_vaccine")

head(covid19_vaccine)
```

### Proportion of vaccinated population

We can measure the proportion of vaccinated group out of the total population using the `people_fully_vaccinated` and `population` fields. We will start by filtering the data by the latest date and removing observations without population data (province level):

```{r}
library(dplyr)

df <- covid19_vaccine %>% 
  filter(date == max(date),
         !is.na(population)) 
```

Next, we will calculate the ratio of the fully vaccinated people out of the total population:

```{r}
df <- df %>% 
  mutate(fully_vaccinated_ratio = people_fully_vaccinated / population) %>%
  arrange(- fully_vaccinated_ratio)

head(df, 10)
```

We can plot the top 20 vaccinated countries:

```{r fig.height=6, fig.width=8}
library(plotly)

top_20 <- df %>% 
  slice_head(n = 20) %>%
  arrange(fully_vaccinated_ratio) %>%
  mutate(country = factor(country_region, levels = country_region))
  
plot_ly(data = top_20,
        y = ~ country,
        x = ~ round(100 * fully_vaccinated_ratio, 2),
        text = ~ paste(round(100 * fully_vaccinated_ratio, 1), "%"),
        textposition = 'auto',
        orientation = "h",
        type = "bar") %>%
  layout(title = "Percentage of Fully Vaccineted Population - Top 20 Countries",
         yaxis = list(title = ""),
         xaxis = list(title = "Percentage", ticksuffix = "%"))

```

### Continent View

Similarly, we can filter and plot the percentage of vaccinated people by continent, using the `continent_name` field:

```{r fig.height=8, fig.width=9}

continent_df <- df %>%
  filter(!is.na(continent_name),
         !is.na(fully_vaccinated_ratio))

table(continent_df$continent_name)

p <- list()

for(i in unique(continent_df$continent_name)){
  d <- NULL
  
  d <- continent_df %>% 
    filter(continent_name == i) %>% 
    arrange(fully_vaccinated_ratio) %>%
    mutate(country = factor(country_region, levels = country_region))
  
  p[[i]] <-  plot_ly(data = d,
          y = ~ country,
          x = ~ round(100 * fully_vaccinated_ratio, 2),
          orientation = "h",
          showlegend = FALSE,
          name = i,
          type = "bar") %>%
    layout(title = "Percentage of Fully Vaccineted Population by Continent and Country",
           yaxis = list(title = ""),
           xaxis = list(title = "", ticksuffix = "%")) %>%
    add_annotations(text = i,
                    xref = "paper",
                    yref = "paper",
                    x = 0.5,
                    y = 0.1,
                    align = "right",
                    showarrow = FALSE)
  
  
}


subplot(p, nrows = 3, shareX = TRUE, margin = 0.06)
  
```
