---
title: "Covid19 Vaccine Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Covid19 Vaccine Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.height=5, fig.width=8, message=FALSE, warning=FALSE,
  collapse = TRUE,
  comment = "#>"
)
```


This vignette provides an overview to the `covid19_vaccine` dataset. 

**Note:** Before getting started, please note for the following:

- It may be challenging to make a comparison between countries as the count method of number of doses may vary from country to country. 
- In addition, the doses summary includes multiple types of vaccines, and there is no information if the doses is a single or multiple doses series. 
- Last but not least, the author of this vignette does not have any domain knowledge in pharma, this is purely exploratory analysis of data

Data sources:

-   Vaccine data - <https://github.com/govex/COVID-19/tree/master/data_tables/vaccine_data/global_data>
-   Country/province codes and population - <https://github.com/CSSEGISandData/COVID-19/blob/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv>
-   Continent names and code - DATA HUB, <https://datahub.io/JohnSnowLabs/country-and-continent-codes-list>


## Load the data

Let's start by loading the packages we will use in this vignette:

```{r}
library(coronavirus)
library(dplyr)
library(plotly)
```

We load the dataset from the **coronavirus** package, and use the **dplyr** and **plotly** packages to manipulate and plot the data. Let's load the data:

```{r}
data("covid19_vaccine")

head(covid19_vaccine)
```


The `covid19_vaccine` dataset provides time-series data on the vaccination progress by country or province (if applicable). Likewise the `coronavirus` dataset, the COVID19 vaccine raw data is collected by Johns Hopkins University Center for Systems Science and Engineering (JHU CCSE). The `covid19_vaccine` data includes the following fields:

- `date` - Data collection date in YYYY-MM-DD format
- `country_region` - Country or region name
- `continent_name` - Continent name
- `continent_code` - Continent code
- `combined_key` - Country and province (if applicable)
-   `doses_admin` - Cumulative number of doses administered. When a vaccine requires multiple doses, each one is counted independently
-   `people_at_least_one_dose` - Cumulative number of people who received at least one vaccine dose. When the person receives a prescribed second dose, it is not counted twice
-   `population` - Country or province population
-   `uid` - Country code
-   `iso2` - Officially assigned country code identifiers with two-letter
-   `iso3` - Officially assigned country code identifiers with three-letter
-   `code3` - UN country code
-   `fips` - Federal Information Processing Standards code that uniquely identifies counties within the USA
-   `lat` - Latitude
-   `long` - Longitude


**Note:** The country / province code fields (e.g., `ios2`, `ios3`, etc.) and population were merged with the raw data


### Refresh the data

While the data in the package is up-to-date with the last date the package was updated on CRAN. That been said, the data is refershed on a daily bases and get update on the package [repository](https://github.com/RamiKrispin/coronavirus), typically there is a one day delay. There are couple of methods to pull the most recent data:

- Use the `update_dataset` which will re-install the package from the package repository. Please see the  vignette for more details 
- Pull the data directly from the [CSV](https://github.com/RamiKrispin/coronavirus/blob/main/csv/covid19_vaccine.csv) file on the repository:

```{r}
library(readr)

vaccine_df <- read_csv(file = "https://raw.githubusercontent.com/RamiKrispin/coronavirus/main/csv/covid19_vaccine.csv",
                       col_types = cols(
                         date = col_date(format = ""),
                         country_region = col_character(),
                         continent_name = col_character(),
                         continent_code = col_character(),
                         combined_key = col_character(),
                         doses_admin = col_number(),
                         people_at_least_one_dose = col_number(),
                         population = col_number(),
                         uid = col_number(),
                         iso2 = col_character(),
                         iso3 = col_character(),
                         code3 = col_number(),
                         fips = col_logical(),
                         lat = col_number(),
                         long = col_number()
                       )) 

# Fixing the continent code field - changing NA to "NA" for North America
vaccine_df$continent_code <- ifelse(is.na(vaccine_df$continent_code) & vaccine_df$continent_name == "North America", "NA", vaccine_df$continent_code)

head(vaccine_df)
```


The plan in the next version is to add similar function to the `refresh_coronavirus_jhu` function to pull the vaccine dataset.

## Analysing the vaccine dataset

The dataset includes two cumulative counts of the number of vaccine doses received - `doses_admin` and `people_at_least_one_dose`. The main difference between the two - the last, `people_at_least_one_dose`, count vaccine required multiple doses per person as one (if the person received the two, otherwise count as one). The first count, `doses_admin`, count each separately. So, for example, if you received the Pfizer or Moderna vaccine, here are it will be count:

- If the person complete the two doses the `people_at_least_one_dose` will count it as one dose and the `doses_admin` will count it as two
- Similarly, if the person received only the first out of two doses and skip the second it will be count in both variables as one

We can make the assumption that the last case is rare and therefore, the `people_at_least_one_dose` could be a good proxy for people that are fully vaccinated. Let's see the differences between the two by using the US vaccine data:


```{r}
covid19_vaccine |> 
  filter(country_region == "US") |>
  select(date, doses_admin, total = people_at_least_one_dose) |>
  arrange(date) |>
  plot_ly() |>
  add_lines(x = ~ date,
            y = ~ doses_admin,
            name = "All Doses") |>
  add_lines(x = ~ date,
            y = ~ total,
            name = "Doses Sets") |>
  layout(title = "US Cumulative Number of Doses by Type",
         legend = list(x = 0.05, y = 0.95),
         margin = list(l = 50, r = 50, b = 50, t = 90),
         yaxis = list(title = "Number of Doses"),
         xaxis = list(title = ""))

```


Here are some thoughts about to leverage this dataset:

- While it is not fully accurate, as mentioned above, the `people_at_least_one_dose` could provides a good proxy for the total number of people that are fully vaccinate. - We can leverage the `doses_admin` column to calculate some stats about the total number of doses provided in a given period or overall. 
- Combining the vaccine and cases datasets together could provide some insights about the vaccine adoption and effective.

### Country level summary

Before going back and diving into the US data, let's summarize the data by country. Since the data is cumulative, let's take a snapshot of the most recent date in the data and calculate the ratio between the total number doses and the size of population:

```{r}
max(covid19_vaccine$date)

df_summary <- covid19_vaccine |>
  filter(date == max(date)) |>
  select(date, country_region, doses_admin, total = people_at_least_one_dose, population, continent_name) |>
  mutate(doses_pop_ratio = doses_admin / population,
         total_pop_ratio = total / population) |>
  arrange(- total)

head(df_summary, 10)

```


It would be interesting to see if the ratio between the number of overall doses received and the size of population. Let's filter missing values and plot the ratio:


```{r}
df_summary |> 
  filter(country_region != "World", 
         !is.na(population),
         !is.na(total)) |>
  plot_ly(x = ~ population,
          y = ~ total,
          text = ~ paste("Country: ", country_region, "<br>",
                         "Population: ", population, "<br>",
                         "Total Doses: ", total, "<br>",
                         "Ratio: ", round(total_pop_ratio, 2), 
                         sep = ""),
          type = "scatter",
          mode = "markers") |>
  layout(title = "Total Doses vs. Population",
         margin = list(l = 50, r = 50, b = 60, t = 70),
         yaxis = list(title = "Number of Doses"),
         xaxis = list(title = "Population Size"))
```

As can see, China and India skewing the plot, we can re-scale the plot with log transformation of the number of doses and population size using the Plotly's built-in log transformation. Let's also color by continent and add diagonal lines to represents the 1 to 1 relationship the median and first  quartile:

```{r}

# Setting the diagonal lines range
line_start <- 10000
line_end <- 1500 * 10 ^ 6

# Filter the data
d <- df_summary |> 
  filter(country_region != "World", 
         !is.na(population),
         !is.na(total)) 


# Replot it
plot_ly() |>
  add_markers(x = d$population,
              y = d$total,
              text = ~ paste("Country: ", d$country_region, "<br>",
                             "Population: ", d$population, "<br>",
                             "Total Doses: ", d$total, "<br>",
                             "Ratio: ", round(d$total_pop_ratio, 2), 
                             sep = ""),
              color = d$continent_name,
              type = "scatter",
              mode = "markers") |>
  add_lines(x = c(line_start, line_end),
            y = c(line_start, line_end),
            showlegend = FALSE,
            line = list(color = "gray", width = 0.5)) |>
  add_lines(x = c(line_start, line_end),
            y = c(0.5 * line_start, 0.5 * line_end),
            showlegend = FALSE,
            line = list(color = "gray", width = 0.5)) |>
  
  add_lines(x = c(line_start, line_end),
            y = c(0.25 * line_start, 0.25 * line_end),
            showlegend = FALSE,
            line = list(color = "gray", width = 0.5)) |>
  add_annotations(text = "1:1",
                  x = log10(line_end * 1.25),
                  y = log10(line_end * 1.25),
                  showarrow = FALSE,
                  textangle = -25,
                  font = list(size = 8),
                  xref = "x",
                  yref = "y") |>
  add_annotations(text = "1:2",
                  x = log10(line_end * 1.25),
                  y = log10(0.5 * line_end * 1.25),
                  showarrow = FALSE,
                  textangle = -25,
                  font = list(size = 8),
                  xref = "x",
                  yref = "y") |>
  add_annotations(text = "1:4",
                  x = log10(line_end * 1.25),
                  y = log10(0.25 * line_end * 1.25),
                  showarrow = FALSE,
                  textangle = -25,
                  font = list(size = 8),
                  xref = "x",
                  yref = "y") |>
  add_annotations(text = "Source: Johns Hopkins University - Centers for Civic Impact",
                  showarrow = FALSE,
                  xref = "paper",
                  yref = "paper",
                  x = -0.05, y = - 0.33) |>
  layout(title = "Covid19 Vaccine - Total Doses vs. Population Ratio (Log Scale)",
         margin = list(l = 50, r = 50, b = 90, t = 70),
         yaxis = list(title = "Number of Doses",
                      type = "log"),
         xaxis = list(title = "Population Size",
                      type = "log"),
         legend = list(x = 0.75, y = 0.05))
```

We can use box plot to plot the total doses and population ration distribution by continent:


```{r}
plot_ly(d,
        y = ~ total_pop_ratio,
        color = ~ continent_name,
        type = "box", boxpoints = "all", jitter = 0.3,
        pointpos = -1.8
        ) |>
  layout(title = "Distribution of Total Doses Admited vs Population Size Ratio",
         margin = list(l = 50, r = 50, b = 60, t = 60),
         yaxis = list(title = "Total Doses/Population Ratio"),
         xaxis = list(title = "Continent")
    
  )
```

### Plotting cases with vaccine data

Let's now return to the number of doses in the US and plot the daily number of new cases with daily number of vaccine doses received. We will use the `people_at_least_one_dose` column as a proxy for number of people that fully vaccinated. First, let's reformat the data from cumulative to daily by subtracting from the cumulative count its first lag:


```{r}
us_vaccine <- covid19_vaccine |> 
  filter(country_region == "US") |>
  select(date, total_doses = people_at_least_one_dose) |>
  mutate(total_doses_lag1 = lag(total_doses, 1),
         daily_doses = total_doses - total_doses_lag1)|>
  select(date, daily_doses) |>
  arrange(date) 


head(us_vaccine)

```

Let's now plot the daily doses:

```{r}
plot_ly(us_vaccine) |>
  add_lines(x = ~ date,
            y = ~ daily_doses)  |>
  layout(title = "US Daily Vaccine Doses Received",
         margin = list(l = 50, r = 50, b = 60, t = 60),
         yaxis = list(title = "Total Doses"),
         xaxis = list(title = ""))
```


One thing to notice from the plot above - since Nov 2022 the data is getting update once in a week. Therefore, you can notice that the daily count is equal to zero and spiked up every 7 days as results of subtracting the data with its first lag. In addition, there is a negative spike on December 1st, most likely due to some data integrity issue on that day. You can empute the value of this day using some estimation method (average of previous and next week). As it is not in the scope of this vignette, let's for now set it as zero: 


```{r}
us_vaccine$daily_doses <- ifelse(us_vaccine$daily_doses < 0, 0, us_vaccine$daily_doses)

plot_ly(us_vaccine) |>
  add_lines(x = ~ date,
            y = ~ daily_doses)  |>
  layout(title = "US Daily Vaccine Doses Received",
         margin = list(l = 50, r = 50, b = 60, t = 60),
         yaxis = list(title = "Total Doses"),
         xaxis = list(title = ""))


```

We can now add the Covid19 cases by loading the dataset using the `refresh_coronavirus_jhu` function, filter it (to US) and merge it with the `us_vaccine` dataset we created above:

```{r}

covid19_cases <- refresh_coronavirus_jhu()

head(covid19_cases)

us_cases <- covid19_cases |>
  filter(location == "US",
         data_type == "cases_new") |>
  select(date, cases = value)


head(us_cases)
```



```{r}
df <- us_vaccine |> 
  left_join(us_cases, by = "date") |>
  select(date, daily_doses, daily_cases = cases) |>
  arrange(date)
head(df)


plot_ly(df) |>
  add_lines(x = ~ date,
            y = ~ daily_cases) |>
  add_lines(x = ~ date,
            y = ~ daily_doses)
```

