---
title: "Covid19 Vaccine Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Covid19R Project Format}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The `covid19_vaccine` dataset provides time-series data on the vaccination progress by country or province (if applicable). Likewise the `coronavirus` dataset, the COVID19 vaccine raw data is collected by Johns Hopkins University Center for Systems Science and Engineering (JHU CCSE). The `covid19_vaccine` data includes the following fields:

 * `country_region` - Country or region name
* `date` - Data collection date in YYYY-MM-DD format
* `doses_admin` - Cumulative number of doses administered. When a vaccine requires multiple doses, each one is counted independently
* `people_partially_vaccinated` - Cumulative number of people who received at least one vaccine dose. When the person receives a prescribed second dose, it is not counted twice
* `people_fully_vaccinated` - Cumulative number of people who received all prescribed doses necessary to be considered fully vaccinated
* `report_date_string` - Data report date in YYYY-MM-DD format
* `uid` - Country code
* `province_state` - Province or state if applicable
* `iso2` - Officially assigned country code identifiers with two-letter
* `iso3` - Officially assigned country code identifiers with three-letter
* `code3` - UN country code
* `fips` - Federal Information Processing Standards code that uniquely identifies counties within the USA
* `lat` - Latitude
* `long` - Longitude
* `combined_key` - Country and province (if applicable)
* `population` - Country or province population

Note that the country / province coding (e.g., `ios2`, `ios3`, etc.) is 
```{r}
library(coronavirus)

data("covid19_vaccine")

head(covid19_vaccine)
```

