---
title: "Covid19 Vaccine Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Covid19 Vaccine Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  fig.height=5, fig.width=8, message=FALSE, warning=FALSE,
  collapse = TRUE,
  comment = "#>"
)
```

The `covid19_vaccine` dataset provides time-series data on the vaccination progress by country or province (if applicable). Likewise the `coronavirus` dataset, the COVID19 vaccine raw data is collected by Johns Hopkins University Center for Systems Science and Engineering (JHU CCSE). The `covid19_vaccine` data includes the following fields:

-   `date` - Data collection date in YYYY-MM-DD format
- `country_region` - Country or region name
- `continent_name` - Continent name
- `continent_code` - Continent code
- `combined_key` - Country and province (if applicable)
- 

-   `doses_admin` - Cumulative number of doses administered. When a vaccine requires multiple doses, each one is counted independently
-   `people_at_least_one_dose` - Cumulative number of people who received at least one vaccine dose. When the person receives a prescribed second dose, it is not counted twice
-   `population` - Country or province population
-   `uid` - Country code
-   `iso2` - Officially assigned country code identifiers with two-letter
-   `iso3` - Officially assigned country code identifiers with three-letter
-   `code3` - UN country code
-   `fips` - Federal Information Processing Standards code that uniquely identifies counties within the USA
-   `lat` - Latitude
-   `long` - Longitude

Data sources:

-   Vaccine data - <https://github.com/govex/COVID-19/tree/master/data_tables/vaccine_data/global_data>
-   Country/province codes and population - <https://github.com/CSSEGISandData/COVID-19/blob/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv>
-   Continent names and code - DATA HUB, <https://datahub.io/JohnSnowLabs/country-and-continent-codes-list>

**Note:** The country / province code fields (e.g., `ios2`, `ios3`, etc.) and population were merged with the raw data

```{r }
library(coronavirus)

data("covid19_vaccine")

head(covid19_vaccine)
```

### Proportion of vaccinated population

We can measure the proportion of the vaccinated group out of the total population using the `people_fully_vaccinated` and `population` fields. We will start by filtering the data by the latest date and removing observations without population data (province-level):

```{r}
library(dplyr)

df <- covid19_vaccine %>% 
  filter(date == max(date),
         !is.na(population)) 
```

Next, we will calculate the ratio of people with at least one dose out of the total population:

```{r}
df <- df %>% 
  mutate(vaccinated_ratio = people_at_least_one_dose / population) %>%
  arrange(- vaccinated_ratio)

head(df, 10)
```

We can plot the top 20 vaccinated countries:

```{r fig.height=6, fig.width=6}
library(plotly)

top_20 <- df %>% 
  slice_head(n = 20) %>%
  arrange(people_at_least_one_dose) %>%
  mutate(country = factor(country_region, levels = country_region))
  
plot_ly(data = top_20,
        y = ~ country,
        x = ~ round(100 * people_at_least_one_dose, 2),
        text = ~ paste(round(100 * people_at_least_one_dose, 1), "%"),
        textposition = 'auto',
        orientation = "h",
        type = "bar") %>%
  layout(title = "Proportion of Population with at Least On Dose- Top 20 Countries",
         yaxis = list(title = ""),
         xaxis = list(title = "Percentage", ticksuffix = "%"))

```

### Continent View

Similarly, we can filter and plot the percentage of vaccinated people by continent, using the `continent_name` field:

```{r fig.height=10, fig.width=11}

continent_df <- df %>%
  filter(!is.na(continent_name),
         !is.na(fully_vaccinated_ratio))

table(continent_df$continent_name)

p <- list()

for(i in unique(continent_df$continent_name)){
  d <- NULL
  
  d <- continent_df %>% 
    filter(continent_name == i) %>% 
    arrange(fully_vaccinated_ratio) %>%
    mutate(country = factor(country_region, levels = country_region))
  
  p[[i]] <-  plot_ly(data = d,
          y = ~ country,
          x = ~ round(100 * fully_vaccinated_ratio, 2),
          orientation = "h",
          showlegend = FALSE,
          name = i,
          type = "bar") %>%
    layout(title = "Percentage of Fully Vaccineted Population by Continent and Country",
           yaxis = list(title = ""),
           xaxis = list(title = "", ticksuffix = "%")) %>%
    add_annotations(text = i,
                    xref = "paper",
                    yref = "paper",
                    x = 0.5,
                    y = 0.1,
                    align = "right",
                    showarrow = FALSE)
  
  
}


subplot(p, nrows = 3, shareX = TRUE, margin = 0.06)
  
```

### Comparing between new cases and vaccination

We can compare the changes in the daily new cases and the cumulative number of fully vaccinated on the country level by merging the `coronavirus` and `covid19_vaccine` datasets.

**Note:** At this point, the `coronavirus` dataset does not have the `ios` country codes. Therefore, a merge between the two datasets may require some manual effort for adjusting the country codes.

In the following example, we will plot the daily number of cases and the total number of fully vaccinated people in the US. We will filter the datasets by country and merge them by date:

```{r}

data("coronavirus")


us_cases <- coronavirus %>% 
  filter(country == "US", 
         type == "confirmed") %>%
  arrange(date) %>%
  select(date, cases) %>%
  left_join(
    covid19_vaccine %>% 
      filter(country_region == "US") %>%
      select(date, people_fully_vaccinated),
    by = "date")

 tail(us_cases)
    
```

As you can see in the table above, the daily number of cases and the aggregate number of fully vaccinated people are not on the same scale. Therefore, we will normalize the two series between 0 and 1:

```{r}

us_cases <- us_cases %>%
  mutate(cases_normalized = (cases - min(cases)) / (max(cases) - min(cases)),
         people_fully_vaccinated_normilized = (people_fully_vaccinated - min(people_fully_vaccinated, na.rm = TRUE)) / 
           (max(people_fully_vaccinated, na.rm = TRUE) - 
              min(people_fully_vaccinated, na.rm = TRUE)))

tail(us_cases)

```

Let's plot the two normalized series together:

```{r}
plot_ly(data = us_cases,
        x = ~ date,
        y = ~ cases_normalized,
        type = "scatter",
        mode = "line",
        name = "Daily Cases (Normalized)") %>%
  add_lines(x = ~ date,
            y = ~ people_fully_vaccinated_normilized,
            name = "Fully Vaccinated - Aggregate (Normalized)") %>%
  layout(title = "US - Daily New Cases vs. Total Vaccinated Population (Normalized)",
         legend = list(orientation = 'h'),
         yaxis = list(title = "Normalized Daily Cases/Total Vaccinated"),
         xaxis = list(title = ""),
         margin = list(b = 60))
```
