---
title: "Geospatial Visualization"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{geospatial_visualization}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  warning = FALSE, message = FALSE,
   fig.height=5, fig.width=8,
  collapse = TRUE,
  comment = "#>"
)
```

The `coronavirus` and `covid19_vaccine` datasets provide country-level information on the COVID-19 cases and vaccination progress, respectively. A common method to communicate and visualize country-level data is with the use of choropleth maps. This vignette focuses on approaches for plotting COVID-19 cases with choropleth maps using the following packages:

-   [**rnaturalearth**](https://docs.ropensci.org/rnaturalearth/) **- P**rovides geo-spatial metadata from the [Natural Earth dataset](https://www.naturalearthdata.com). More details available here

-   [**sf**](https://r-spatial.github.io/sf/) **-** A package that provides [simple features access](https://en.wikipedia.org/wiki/Simple_Features) for R

-   [**mapview**](https://r-spatial.github.io/mapview/) **-** A wrapper for the leaflet library

-   [**tmap**](https://r-tmap.github.io/tmap/) **-** A package for creating a thematic maps

-   [**ggplot2**](https://ggplot2.tidyverse.org/) **-** Is a system for declaratively creating graphics

-   [**viridis**](https://sjmgarnier.github.io/viridis/) **-** A package that provide a series of color maps

```{r libraries}
library(rnaturalearth)
library(sf)
library(mapview)
library(tmap)
library(ggplot2)
library(viridis)
```

**Note:** This vignette is not available on the CRAN version (due to size limitations). Therefore, as the packages above are not on the dependencies list of the **coronavirus** package, you may need to install them before.


**Additional setting:** following changes in the default options of the **sf** package from version `1.0-1` by default option is to use s2 spherical geometry as default when coordinates are ellipsoidal. That causes some issues with the **tmap** package, therefore we will set this functionality as `FALSE`:

``` {r set_s2}
sf_use_s2(FALSE)
```

More details are available on this [issue](https://github.com/RamiKrispin/coronavirus/issues/95) and follow-up [issue](https://github.com/r-spatial/sf/issues/1821).

## Data prep

Let's get started by loading the data:

```{r data}
library(coronavirus)

data("covid19_vaccine")

head(covid19_vaccine)
```

We will use the `ne_countries` function from the **rnaturalearth** package to pull the country geometric data:

```{r}
library(dplyr)

map <- ne_countries(returnclass = "sf") %>%
  dplyr::select(name, iso2 = iso_a2, iso3 = iso_a3, geometry)

head(map)


```

```{r}

df <- map %>% left_join(
   covid19_vaccine %>%
    filter(date == max(date)) %>%
    mutate(perc = round(100 * people_at_least_one_dose / population, 2)) %>%
    select(country_region, iso2, iso3, people_at_least_one_dose, perc, continent_name),
    by = c("iso2", "iso3")
)

class(df)
```

After we merge the country data with the corresponding geometry data, it is straightforward to plot the data as `sf` object.

## Choropleth maps with the mapview package

The **mapview** package, a wrapper for the leaflet library, enables to plot `sf` objects seamlessly. Let's start by plotting the percentage of the population that fully vaccinated by country using the `perc` variable:

```{r}
df  %>%
  mapview::mapview(zcol = "perc")
```

By default, the function uses a continuous color scale for the objects (in this case, percentage of population that is vaccinated) color. We can modify it and set color buckets by using the `at` argument. Also, we can define the legend title with the use of the `layer.name` argument:

```{r}
df  %>%
  mapview::mapview(zcol = "perc", 
                   at = seq(0, max(df$perc, na.rm = TRUE), 10), 
                   legend = TRUE,
                   layer.name = "Fully Vaccinated %")
```

Some of the missing values in the plot are un-populated areas such as Antarctica or a territory counted under different states such as Greenland. We can remove those and re-plot the map:

```{r}
df1 <- df %>% 
  filter(!name %in% c("Greenland", "Antarctica"))

df1  %>%
  mapview::mapview(zcol = "perc", 
                   at = seq(0, max(df1$perc, na.rm = TRUE), 10), 
                   legend = TRUE,
                   layer.name = "Fully Vaccinated %")
```

Last but not least, you can customize the color palette using the `col.regions` argument. In the following example, we will use the `plasma` function from the **viridis** package:

```{r}

df1  %>%
  mapview::mapview(zcol = "perc", 
                   at = seq(0, max(df1$perc, na.rm = TRUE), 10), 
                   legend = TRUE,
                   col.regions = plasma(n = 8),
                   layer.name = "Fully Vaccinated %")

```


## Choropleth maps with the tmap package

The **tmap** (thematic maps) package is another useful tool for creating choropleth maps. Similarly to the **mapview** package, the **tmap** package supports `sf` objects. This **tmap** follows the **ggplot2** package syntax style (e.g., using the `+` symbol to add to the plot additional plot). We will use the `tm_shape` function to define the input object with the `df1` object, and add the `tm_polygons` function to customize the plot:

```{r}

tm_shape(df1) + 
  tm_polygons(col = "perc",  
              n = 8,
              title = "Fully Vaccinated %",
              palette = "Blues")
```

The `projection` argument enables setting the map projections method using `PROJ.4` code:

```{r}

tm_shape(df1) + 
  tm_polygons(col = "perc",  
              n = 8,
              projection = 3857,
              title = "Fully Vaccinated %",
              palette = "Blues") 
```

We can use the `continent_name` column to filter and plot the data by specific continent. For example, let's plot the countries in South America:

```{r}
df2 <- df1 %>%
  filter(continent_name == "South America")


tm_shape(df2) + 
  tm_polygons(col = "perc",  
              n = 5,
              title = "Perc. Group",
              palette = "Blues") 

```

Let's customize the plot and use the `tm_stype` function to set the background and the `tm_layout` function to add the plot title:

```{r}
tm_shape(df2) + 
  tm_polygons(col = "perc",  
              n = 5,
              title = "Perc. Group",
              palette = "Blues") + 
  tm_style("cobalt") + 
  tm_text("iso3", size = 0.7) +
  tm_layout(
            title= "% of Population Fully Vaccinated", 
            title.position = c('right', 'top') ,
            inner.margins = c(0.02, .02, .1, .25))
```

You can map the output interactive by setting the `tmap_mode` function to a `view` mode:

```{r}
tmap_mode("view")

tm_shape(df2) + 
  tm_polygons(col = "perc",  
              n = 5,
              title = "Perc. Group",
              palette = "Blues") 
```

Last but not least, we can add facets with the use of the `tm_facets` and `tmap_options` functions:

```{r}

tmap_mode("plot")


tm_shape(df2) + 
  tm_polygons(col = "perc",  
              n = 5,
              title = "Perc. Group",
              palette = "Greens") + 
  tmap_options(limits = c(facets.view = 13)) + 
  tm_facets(by = "name") 
  
```

## Choropleth maps with the ggplot2 package

The `geom_sf` function enables to read and plot `sf` objects with **ggplot2**, for example:

```{r}

ggplot(data = df2, aes(fill = `perc`)) + 
  geom_sf() + 
  scale_fill_viridis_b()

```

The `scale_fill_viridis_b` defines the color setting. By default, the y and x axes represent the earth's coordinates. We can remove those coordinates with the `theme_void` function and customize the color palette with the `scale_fill_viridis` function:

```{r}
ggplot(data = df2, aes(fill = `perc`)) + 
  geom_sf(size = 0.1) + 
    scale_fill_viridis(alpha = 0.9,
                       begin = 0.01,
                       discrete = FALSE,
                       end = 0.9) + 
  geom_sf_label(aes(label = name)) + 
  labs(fill = "Percentage",
       title = "'% of Population Fully Vaccinated",
       caption = "Source: Johns Hopkins University Center for Systems Science and Engineering") + 
   theme_void()

```

#### Sources:

* **mapview** package documentation - https://r-spatial.github.io/mapview/
* **tmap** package documentation - https://r-tmap.github.io/tmap/
* **ggplot2** package documentation - https://ggplot2.tidyverse.org/
* `geom_sf` function - https://ggplot2.tidyverse.org/reference/ggsf.html

