---
title: "Introduction to the Coronavirus Dataset"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to the Coronavirus Dataset}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message=FALSE, 
                      warning=FALSE, 
                      fig.height=5, 
                      fig.width=8,
                      collapse = TRUE,
                      comment = "#>")
```

## The coronavirus dataset

The `coronavirus` dataset provides a daily summary of COVID-19 cases by geographic location (i.e., country/province). It includes total daily confirmed, death, and recovery^1^ cases. Let’s load the dataset from the **coronavirus** package:


```{r }
library(coronavirus)

data(coronavirus)
```



The dataset has the following fields:

* `date` - The date of the summary
* `province` - The province or state, when applicable
* `country` - The country or region name
* `Lat` - Latitude point
* `Long` - Longitude point
* `type` - the type of case (i.e., confirmed, death)
* `cases` - the number of daily cases (corresponding to the case type)
* `uid` - Country code
* `iso2` - Officially assigned country code identifiers with two-letter
* `iso3` - Officially assigned country code identifiers with there-letter
* `code3` - UN country code
* `combined_key` - Country and province (if applicable)
* `population` - Country or province population
* `continent_name` - Continent name
* `continent_code` - Continent code


We can use the `head` and `str` functions to see the structure of the dataset:

```{r }
head(coronavirus)

str(coronavirus)
```


^1^ Recovery data is discontinued from Aug 5th, please see the following [issue](https://github.com/RamiKrispin/coronavirus/issues/71) for more details.


## Querying and analyzing the coronavirus dataset

In the example below, we will use the **dplyr** and **tidyr** packages to query, transform, reshape, and keep the data tidy, the **plotly** package to plot the data, and the **DT** package to view it:

```{r}
library(dplyr)
library(tidyr)
library(plotly)
library(DT)
```

### Cases summary

Let's start with summarizing the total number of cases by type as of `r max(coronavirus$date)` and then plot it: 

```{r}
total_cases <- coronavirus %>%
  filter(type != "recovery") %>%
  group_by(type) %>%
  summarise(cases = sum(cases)) %>%
  mutate(type = factor(type, levels = c("confirmed", "death"))) 

total_cases
```


Likewise, we can summarise the data by continent using the `continent_name` field:

```{r}
coronavirus %>%
  filter(type != "recovery") %>%
  group_by(type, continent_name) %>%
  summarise(cases = sum(cases), .groups = "drop") %>%
  mutate(type = factor(type, levels = c("confirmed", "death"))) %>%
  pivot_wider(names_from = type, values_from = cases) %>%
  mutate(death_rate = death / confirmed) %>%
  filter(!is.na(continent_name)) %>%
  arrange(-death_rate) %>%
  datatable(rownames = FALSE,
            colnames = c("Continent", "Confrimed Cases", "Death Cases","Death Rate %")) %>%
  formatPercentage("death_rate", 2)
```

 
You can use those numbers to derive the current worldwide death rate (percentage):

```{r}
round(100 * total_cases$cases[2] / total_cases$cases[1], 2)
```

### Worldwide aggregation 

Let's group the data by the `date` and case `type` fields and aggregate the total cases (confirmed and death cases) to the worldwide level and plot the two side by side:

```{r}
df <- coronavirus %>%
  filter(type != "recovery") %>%
  group_by(date,type) %>%
  summarise(total = sum(cases), .groups = "drop")

p_1 <- plot_ly(data = df %>% filter(type == "confirmed"),
        x = ~ date,
        y = ~ total,
        name = "Confirmed",
        type = "scatter",
        mode = "line") %>%
  layout(yaxis = list(title = "Cases"),
         xaxis = list(title = ""))

p_2 <- plot_ly(data = df %>% filter(type == "death"),
              x = ~ date,
              y = ~ total,
              name = "Death",
              line = list(color = "red"),
              type = "scatter",
              mode = "line") %>%
  layout(yaxis = list(title = "Cases"),
         xaxis = list(title = "Source: Johns Hopkins University Center for Systems Science and Engineering"))

subplot(p_1, p_2, nrows = 2, 
              titleX = TRUE,
              titleY = TRUE) %>%
  layout(title = "Worldwide - Daily Confirmed and Death Cases",
         margin = list(t = 60, b = 60, l = 40, r = 40),
         legend = list(x = 0.05, y = 1)
         )
```



### Top effected countries

The next table provides an overview of the ten countries with the highest confirmed cases. We will use the `datatable` function from the **DT** package to view the table:


```{r}
confirmed_country <- coronavirus %>% 
  filter(type == "confirmed") %>%
  group_by(country) %>%
  summarise(total_cases = sum(cases)) %>%
  mutate(perc = total_cases / sum(total_cases)) %>%
  arrange(-total_cases)

confirmed_country %>%
  head(10) %>%
  datatable(rownames = FALSE,
            colnames = c("Country", "Cases", "Perc of Total")) %>%
  formatPercentage("perc", 2)
```


The next plot summarize the distribution of confirmed cases by country:

```{r}
conf_df <- coronavirus %>% 
  filter(type == "confirmed") %>%
  group_by(country) %>%
  summarise(total_cases = sum(cases)) %>%
  arrange(-total_cases) %>%
  mutate(parents = "Confirmed") %>%
  ungroup() 
  
  plot_ly(data = conf_df,
          type= "treemap",
          values = ~total_cases,
          labels= ~ country,
          parents=  ~parents,
          domain = list(column=0),
          name = "Confirmed",
          textinfo="label+value+percent parent")
```

### Death rates

Similarly, we can use the `pivot_wider` function from the **tidyr** package (in addition to the **dplyr** functions we used above) to get an overview of the confirmed and death cases and calculate the death rates:

```{r}
coronavirus %>% 
  filter(country != "Others") %>%
  group_by(country, type) %>%
  summarise(total_cases = sum(cases)) %>%
  pivot_wider(names_from = type, values_from = total_cases) %>%
  arrange(- confirmed) %>%
  mutate(death_rate = death / confirmed)  %>%
  datatable(rownames = FALSE,
            colnames = c("Country", "Confirmed","Death", "Death Rate")) %>%
   formatPercentage("death_rate", 2) 
```
