---
title: "Introduction to the Coronavirus Dataset"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to the Coronavirus Dataset}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message=FALSE, 
                      warning=FALSE, 
                      fig.height=5, 
                      fig.width=8,
                      collapse = TRUE,
                      comment = "#>")
```

### The coronavirus dataset

The `coronavirus` dataset provides a snapshot of the daily confirmed, recovered, and death cases of the Coronavirus (the 2019 Novel Coronavirus COVID-19) by geographic location (i.e., country/province). Let's load the dataset from the **coronavirus** package:

```{r }
library(coronavirus)

data(coronavirus)
```



The dataset has the following fields:

* `date` - The date of the summary
* `Province.State` - The province or state, when applicable
* `Country.Region` - The country or region name
* `Lat` - Latitude point
* `Long`- Longitude point
* `cases` number of daily cases (i.e., confirmed, death, and recovered)

We can use the `head` and `str` functions to see the structure of the dataset:

```{r }
head(coronavirus)

str(coronavirus)
```
