---
title: "Introduction to the Coronavirus Dataset"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to the Coronavirus Dataset}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      message=FALSE, 
                      warning=FALSE, 
                      fig.height=5, 
                      fig.width=8,
                      collapse = TRUE,
                      comment = "#>")
```

### The coronavirus dataset

The `coronavirus` dataset provides a snapshot of the daily confirmed, recovered, and death cases of the Coronavirus (the 2019 Novel Coronavirus COVID-19) by geographic location (i.e., country/province). Let's load the dataset from the **coronavirus** package:

```{r }
library(coronavirus)

data(coronavirus)
```



The dataset has the following fields:

* `date` - The date of the summary
* `Province.State` - The province or state, when applicable
* `Country.Region` - The country or region name
* `Lat` - Latitude point
* `Long`- Longitude point
* `cases` number of daily cases (i.e., confirmed, death, and recovered)

We can use the `head` and `str` functions to see the structure of the dataset:

```{r }
head(coronavirus)

str(coronavirus)
```

### Querying and analyzing the coronavirus dataset

We will use the **dplyr** and **tidyr** packages to query, transform, reshape, and keep the data tidy, the **plotly** package to plot the data and the **DT** package to view it:

```{r}
library(dplyr)
library(tidyr)
library(plotly)
library(DT)
```

#### Cases summary

Let's start with summarizing the total number of cases by type as of `r max(coronavirus$date)` and then plot it: 

```{r}
total_cases <- coronavirus %>% 
  group_by(type) %>%
  summarise(cases = sum(cases)) %>%
  mutate(type = factor(type, levels = c("confirmed", "recovered", "death")))

total_cases
```


```{r}
plot_ly(data = total_cases, 
        x = ~ type, 
        y = ~cases, 
        type = 'bar',
        text = ~ paste(type, cases, sep = ": "),
    hoverinfo = 'text') %>%
  layout(title = "Coronavirus - Cases Distribution",
         yaxis = list(title = "Number of Cases"),
         xaxis = list(title = "Case Type"),
         hovermode = "compare")
```


We can learn from this table that, so far worldwide, the recovery rate is `r paste(round(100 *total_cases$cases[3] / total_cases$cases[1], 2), "%", sep = "")` and the death rate is `r paste(round(100 *total_cases$cases[2] / total_cases$cases[1], 2), "%", sep = "")`.

#### Top effected countries

The next table provides an overview of the ten countries with the highest confirmed cases. We will use the `datatable` function from the **DT** package to view the table:


```{r}
confirmed_country <- coronavirus %>% 
  filter(type == "confirmed") %>%
  group_by(Country.Region) %>%
  summarise(total_cases = sum(cases)) %>%
  mutate(perc = total_cases / sum(total_cases)) %>%
  arrange(-total_cases)

confirmed_country %>%
  head(10) %>%
  datatable(rownames = FALSE,
            colnames = c("Country", "Cases", "Perc of Total")) %>%
  formatPercentage("perc", 2)
```


As `r paste(round(100 * confirmed_country$perc[which(confirmed_country$Country.Region == "Mainland China")],1), "%", sep = "")` of the confirmed cases are in China, to get a better overview of the worldwide dist of the virus, let's exclude China and plot the rest of the world dist:

```{r}
coronavirus %>% 
  filter(type == "confirmed", 
         Country.Region != "Mainland China") %>%
  group_by(Country.Region) %>%
  summarise(total_cases = sum(cases)) %>%
  arrange(-total_cases) %>%
  mutate(country = factor(Country.Region, levels = Country.Region)) %>%
  ungroup() %>%
  plot_ly(labels = ~ country, 
          values = ~ total_cases,
          type = "pie",
          textposition = 'inside',
        textinfo = 'label+percent',
        insidetextfont = list(color = '#FFFFFF'),
        hoverinfo = 'text',
        text = ~ paste(country, "<br />",
                     "Number of confirmed cases: ", total_cases, sep = "")) %>%
  layout(title = "Coronavirus - Confirmed Cases (Excluding China)")
```
