---
title: "Showing the Spatial Distribution of Covid-19 Confirmed Cases"
author: "Jarrett Byrnes"
date: "3/11/2020"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Put the title of your vignette here}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
library(ggplot2)

```

## Updating the coronavirus data

To see the current distribution of cases around the world, we'll start by updating the coronavirus dataset.

```{r update, results='hide'}
library(coronavirus)
library(dplyr)

```

## Generating spatial data

The `coronavirus` package using the [rnaturalearth](https://cran.r-project.org/web/packages/rnaturalearth/) package for spatial information. [Natural Earth](https://www.naturalearthdata.com/) provides a wealth of spatial data easily accessed via R as [sf](https://cran.r-project.org/web/packages/sf/) objects. Using [sf](https://cran.r-project.org/web/packages/sf/), we can do a spatial join on the coronavirus data to get it synced up with the `ne_countries()` data from `rnaturalearth`.

Let's get the data as both points for plotting as well as polygons we can fill. We will filter to confirmed cases only.

```{r make_sf_objs}
library(sf)
library(rnaturalearth)

coronavirus_points <- coronavirus_spatial() %>%
                       filter(type == "confirmed")
                       
coronavirus_polys <- coronavirus_spatial(return_shape = "polygon")%>%
                       filter(type == "confirmed")
```

## Aggregating data

Let's aggregate so that we only look at total confirmed cases.

```{r aggregate}
library(dplyr)
coronavirus_points <- coronavirus_points %>%
  group_by(Province.State, Country.Region, name, continent) %>%
  summarize(cases = sum(cases))

coronavirus_polys <- coronavirus_polys %>%
  group_by(Country.Region, name, continent) %>%
  summarize(cases = sum(cases))
```

## Plotting

Great! We can now see what the spatial distribution looks like simply using ggplot. Note, the polys do not contain any records with nothing in them, so we'll need a baseline worldmap.

```{r plot1, fig.width=6}
library(ggplot2)

worldmap <- ne_countries(returnclass = "sf")

ggplot(data = worldmap) +
  geom_sf(fill = "white") +
  geom_sf(data = coronavirus_polys,
          mapping = aes(fill = log10(cases+1))) +
  geom_sf(data = coronavirus_points,
          mapping = aes(size = cases),
          alpha = 0.7, color = "black") +
  scale_fill_viridis_c(option = "D",
                       breaks = 0:4, labels = 10^c(0:4)) +
  scale_size_continuous(range = c(1, 8)) +
  labs(fill = "# of Cases", size = "# of Cases") +
  theme_minimal()
```

Note, `rnaturalearth` returns lots of information, such as population size, which we might want to get percentages, etc.

## Single Country
What if we had wanted to just do a single country? We can manually implement the steps inside of the `coronavirus_spatial` ourselves quite easily. Let's look at the continental US, for example. Rather than doing a lot of string manipulation on the `Province.State` column, we can use `rnaturalearth` with an `st_join` to get the info we want for grouping calculations.

```{r make_usa}
usa <- ne_states(country = "United States of America", returnclass = "sf")

usa_confirmed_sf <- coronavirus %>%
  filter(Country.Region=="US") %>%
  filter(type=="confirmed") %>%
  st_as_sf(coords = c(x = "Long", y = "Lat"),
           crs = st_crs(usa)) 

usa_points <- st_join(usa_confirmed_sf, usa) %>%
  group_by(name) %>%
  summarize(cases = sum(cases)) %>%
  filter(!(name %in% c("Hawaii", "Alaska"))) %>%
  filter(!is.na(name))

usa_polygons <- st_join(usa, usa_confirmed_sf) %>%
  group_by(name) %>%
  summarize(cases = sum(cases))%>%
  filter(!(name %in% c("Hawaii", "Alaska")))%>%
  filter(!is.na(name))
```

We can now see how things are as of the data downloaded on `r format(Sys.time(), "%a %b %d %Y")`.

```{r plot_us_cases, fig.width=6}
ggplot(data = usa %>% filter(!(name %in% c("Hawaii", "Alaska")))) +
  geom_sf(fill = "white") +
  geom_sf(data = usa_points, mapping = aes(size = cases),
          alpha = 0.7) +
  theme_minimal()
```

Or we can look at things state by state using fill.

```{r plot_us_cases_poly, fig.width=6}
ggplot(data = usa %>% filter(!(name %in% c("Hawaii", "Alaska")))) +
  geom_sf(fill = "white") +
  geom_sf(data = usa_polygons, mapping = aes(fill = cases)) +
  theme_minimal() +
  scale_fill_viridis_c(na.value = "white")
```


