---
title: "Update the coronavirus Dataset"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Update the coronavirus Dataset}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

While the CRAN version of the package is updated once every month or two, the Github (Dev) version is updating on a daily bases. The following options allow you to keep the data updated with the ones available on the Dev version:

* Use the `update_dataset` function
* Read directly from the package repo a csv format of the data

### The update_dataset function

The `update_dataset` function enables to keep the installed version updated with the data available on [Github](https://github.com/RamiKrispin/coronavirus). The function compared between the dataset on the installed version and the ones on the Dev version:

``` r 
library(coronavirus)

update_dataset()
```

If no new data is available on the Dev version, the function will return the following message:

``` r
No updates are available
```

Once new data is available, the function will prompt the following question that enables the user to select whether to install the updates from the dev version:

``` r
Updates are available on the coronavirus Dev version, do you want to update? n/Y
```

In order to make the new data available, you will have to restart your R session.

**Note:** As frequent changes may occur on the raw data structure (such as new fields, retroactive updates in the data, etc.), the Dev version dataset may change accordingly. 


### Reading the data from CSV version

Alternatively, you can read and load the data directly from the package repository, using the csv version:

```{r}
library(readr)
coronavirus_df <- read_csv("https://raw.githubusercontent.com/RamiKrispin/coronavirus/master/csv/coronavirus.csv", 
    col_types = cols(date = col_date(format = "%Y-%m-%d"), 
        cases = col_number()))

head(coronavirus_df)
```

The main difference between the first method (the `update_dataset` function) and the second method (reading a CSV format of the data) is that the date field on the last method is not formated as a `Date` object. A quick reformating can fix it:

```{r}
str(coronavirus_df)
```


